Likelihoods for a general class of ARGs under the SMC

### 2025-05-11 paper.tex
add pseudocode

@@

This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see Fig.~\ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along
the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in Eq.~\eqref{eq:full-lik} using $I$, and then update the counts in $I$.
We further keep track of
the last parent a child node was connected to to detect recombination events.

@@

>>>>

% For algorithms in supplementary materials
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}

@@

This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see Fig.~\ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along
the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in Eq.~\eqref{eq:full-lik} using $I$, and then update the counts in $I$.
We further keep track of
the last parent a child node was connected to to detect recombination events.
The pseudocode for the algorithm is included in Alg.~\ref{alg:full}. Note that Alg.~\ref{alg:depth} 
details how to compute the integrals contained in Eq.~\eqref{eq:depth}.

@@

% algorithm in pseudocode
%% Algorithm to compute contribution of time component of each edge

\begin{algorithm}[H]
\caption{\textsc{LogDepth}: pseudocode for equation Eq.~\eqref{eq:depth}}
\label{alg:depth}
\KwIn{
    Count vector $C$; interval boundaries $T$;\\
    Parent time $t_p$; child time $t_c$;\\
    Initial index $i_0$; recombination rate $r$;\\
    Coalescence rate $c$; recombination flag $R$
}
\KwOut{log likelihood contribution in time dimension}
\BlankLine

Initialize cumulative area $A \leftarrow 0$\;
Initialize result $S \leftarrow 0$\;
Set index $i \leftarrow i_0$\;

Define helper function $f(f_0, f_1)$:
\begin{itemize}
    \item If $f_0 = f_1$, return $0$
    \item Else return $\displaystyle \frac{c}{r - c f_1} \cdot \frac{c}{r - c f_0} \cdot (f_1 - f_0) \cdot \frac{r}{c}$
\end{itemize}

Set stopping time:
\[
t_{\text{stop}} \leftarrow
\begin{cases}
t_c & \text{if } R = \text{False} \\
t_p & \text{otherwise}
\end{cases}
\]

\While{$T_i > t_{\text{stop}}$}{
    $A \leftarrow A + (T_i - T_{i-1}) \cdot C_{i-1}$\;
    $i \leftarrow i - 1$\;
    \If{$i < 1$}{break}
}

\If{$R = \text{False}$}{
    \Return $-c \cdot A$
}

\[
B \leftarrow - \frac{c}{r - c C_{i-1}} \cdot \frac{r}{c} \cdot \exp(-r T_i - c A)
\]

\While{$i > 0$}{
    $A \leftarrow A + (T_i - T_{i-1}) \cdot C_{i-1}$\;
    $i \leftarrow i - 1$\;
    \If{$T_i = t_c$}{break}
    \[
    S \leftarrow S + f(C_{i-1}, C_i) \cdot \exp(-r T_i - c A)
    \]
}

\[
S \leftarrow S + \frac{c}{r - c C_i} \cdot \frac{r}{c} \cdot \exp(-r T_i - c A)
\]

\[
S \leftarrow S + B
\]

\Return $\log(S)$

\end{algorithm}

%% Main algorithm

\begin{algorithm}[H]
\caption{\textsc{LogLikelihood}: pseudocode for Eq.~\eqref{eq:full-lik}}
\label{alg:full}
\KwIn{
    Time vector $I$ for all nodes;\\
    Edge arrays containing parent, child, left and right coordinate of each edge: \texttt{edges\_parent}, \texttt{edges\_child}, \texttt{edges\_left}, \texttt{edges\_right};\\
    Recombination rate $r$, coalescence rate $c$;\\
    \texttt{treesequence};\\
}
\KwOut{Total log-likelihood value}
\BlankLine

Initialize log-likelihood: $\text{LL} \leftarrow 0$ \;
Initialize lineage count vector $C \leftarrow 0$ \;
Initialize arrays to track last parent and visited nodes \;

\While{next tree exists in treesequence}{

    \tcp{Process edge removals}
    \ForEach{edge $e$ being removed}{
        Identify parent $p$ and child $c$ from $e$ \;
        Get node times $t_p$ for $p$, and $t_c$ for $c$ \;
        Decrease lineage count $C$ between $t_c$ and $t_p$ \;
        Record $p$ as last parent of $c$ \;
    }

    \tcp{Process edge insertions}
    \ForEach{edge $e$ being inserted}{
        Identify parent $p$ and child $c$ from $e$ \;
        Get node times $t_p$ for $p$, and $t_c$ for $c$ \;

        Set \texttt{rec\_event} $\leftarrow$ False \;
        Set \texttt{left\_parent\_time} $\leftarrow \infty$ \;

        \If{child $c$ had a previous parent $p^{\prime}$}{
            Retrieve previous parent time $t_{p^{\prime}}$\;
            \If{$p^{\prime}$ $\neq p$ and $p$ not visited}{
                Set \texttt{rec\_event} $\leftarrow$ True \;
            }
        }

        Set $t_{p_{\min}} \leftarrow \min(t_p, \texttt{left\_parent\_time})$ as effective parent time \;

        \tcp{Add depth contribution}
        $\text{LL} \leftarrow \text{LL} + \textsc{LogDepth}(C, I, t_{p_{\min}}, t_c, p, r, c, \texttt{rec\_event})$

        \tcp{Add span contribution}
        $\text{LL} \leftarrow \text{LL} -r \cdot (t_p - t_c) \cdot (\texttt{edges\_right}[e] - \texttt{edges\_left}[e])$

        \tcp{Update lineage count}
        Increase $C$ between $t_c$ and $t_p$ \;
    }
}

Number of nodes with out-degree 0: $n_o$ \;
Contribution coalescent events: $ll_c = \text{num\_edges} -\text{num\_nodes} + n_o$ \; 
Number of nodes with in-degree 0: $n_r$ \;
Contribution recombination events: $ll_r = \text{num\_edges} - \text{num\_nodes} + n_r$ \;
Update likelihood: $\text{LL} \leftarrow \text{LL} +ll_c \cdot \log(c) + ll_r \cdot \log(r)$ \;

\Return $\text{LL}$

\end{algorithm}
